---
title: "Prior Data"
author: "BPR"
date: "2023-05-18"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Initialize Matrix
```{r readData}
load(file.choose())                         # Hurricane "NISData.RData"
NIS_Data -> data_df
rm(list = "NIS_Data")
```
These data are NOT from the PCL-5, so finding a clinicial cut-off for PTSD is arbitrary. Still, let's try it.


```{r steadyState}
# Steady state
4 -> cutoff
matrix(c(length(which(data_df$PTSD < cutoff)),
         length(which(data_df$PTSD >= cutoff)))) ->
  ss
ss / sum(ss) ->
  ss
```
Now, we have the following two equations:
$(a-\lambda)s_1 + (1-a)s_2 = 0$
$(1-d)s_1 + (d-\lambda)s_2) = 0$

We want d > 0.5 (those in distress tend to stay in distress), and we know $s_1$ and $s_2$, so we solve the above for *a* and $\lambda$ numerically. After rearranging, we get:
$a(s_1 - s_2) + \lambda(-s_1) = -s_2$
$\lambda(-s_2) = (d-1)s_1 -ds_2$

```{r initializeMarkov}
# Prior for transitions
0.7 -> d                                    # Fraction who remain distressed
solve(a = matrix(c(ss[1] - ss[2],
                   -ss[2],
                   0,
                   -ss[2]),
                 ncol = 2,
                 nrow = 2,
                 byrow = TRUE),
      b = c(-ss[2],
            (d - 1) * ss[1] - d * ss[2])) ->
  al

matrix(c(al[1], 1 - al[1],
         1-d, d),
       ncol = 2,
       nrow = 2,
       byrow = TRUE) ->
  M
```

# Updating
Updating goes with two things:

1. A state change could be reported
2. The bounds on the entries of M could indicate a significant change



# Second Data Set: RE-WIRE
```{r}
library(tidyverse)
library(haven)
# Start with the REWIRE imputed data
read_spss(file.choose()) %>%
  select(starts_with("pcl5")) ->
  pcl_df
paste0(colnames(pcl_df)[1:20],
       "_s1") ->
  colnames(pcl_df)[1:20]
zap_label(pcl_df) -> pcl_df
zap_labels(pcl_df) -> pcl_df
zap_formats(pcl_df) -> pcl_df

pcl_df %>%
  mutate(PCL1 = rowSums(across(ends_with("_s1")))) %>%
  mutate(PCL2 = rowSums(across(ends_with("_s2")))) %>%
  mutate(PCL3 = rowSums(across(ends_with("_s3")))) %>%
  mutate(PCL4 = rowSums(across(ends_with("_s4")))) %>%
  mutate(PCL5 = rowSums(across(ends_with("_s5")))) %>%
  mutate(PCL6 = rowSums(across(ends_with("_s6")))) ->
  dum1
```

